---
title: "Getting Started: Uncensored Continuous Treatments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: Uncensored Continuous Treatments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
set.seed(42)
```

## Overview

This vignette walks through the core `copulagps` workflow for estimating
causal effects of two correlated continuous treatments in an observational
study.  We simulate data from a known data-generating process so that we
can verify the method recovers the true ATE.

## Simulate data

We generate n = 500 observations with two treatments drawn from a bivariate
Clayton copula (Kendall τ = 0.5) and Normal marginals that depend on five
covariates.

```{r simulate}
library(copulagps)

n  <- 500
d  <- 2

# Covariates: 3 continuous + 2 binary
X <- cbind(
  X1 = rnorm(n), X2 = rnorm(n), X3 = rnorm(n),
  X4 = rbinom(n, 1, 0.5), X5 = rbinom(n, 1, 0.5)
)

sim <- simulate_treatments(
  n             = n,
  copula_family = "clayton",
  tau           = 0.5,
  covariate_mat = X,
  beta_list     = list(c(0.4, 0, 0, 0.4, 0), c(0.4, 0, 0, -0.2, 0))
)
T_mat <- sim$y   # n x 2 treatment matrix

# True ATE for the additive outcome: Delta = t_h1 - t_l1 + t_h2 - t_l2
# We'll compute it exactly after we choose the contrast.

# Helper to generate outcomes (reused across vignettes)
sim_outcome <- function(T_mat, X_mat, binary = FALSE, interaction = FALSE) {
  lin <- T_mat[, 1] + T_mat[, 2] + 0.5 * X_mat[, 1]
  if (interaction) lin <- lin + 0.5 * T_mat[, 1] * T_mat[, 2]
  if (binary) {
    rbinom(nrow(T_mat), 1, plogis(-3 + lin))
  } else {
    lin + rnorm(nrow(T_mat), 0, 1.5)
  }
}

Y_cont   <- sim_outcome(T_mat, X, binary = FALSE)
Y_binary <- sim_outcome(T_mat, X, binary = TRUE)
```

## Estimate the GPS

```{r gps}
gps <- estimate_gps(
  y             = T_mat,
  x             = X,
  copula_family = "auto",   # AIC selects among Clayton/Gumbel/Frank/Gaussian
  trim_method   = "hull"    # convex-hull trimming for d = 2
)
print(gps)
```

## Assess covariate balance

```{r balance, fig.alt = "MSMD bar chart before and after GPS weighting"}
bal <- balance_assessment(gps, covariates = X)
print(bal)
plot(bal)
```

Values of MSMD below 0.1 after weighting indicate that GPS reweighting has
removed the association between the treatments and covariates.

## Estimate the ATE — continuous outcome

```{r ate_cont}
# Treatment contrast: 75th vs 25th percentile (default)
ate_cont <- estimate_ate(gps, outcome = Y_cont, covariates = X)
print(ate_cont)
get_ate_table(ate_cont)
```

The true ATE is `d = 2` (sum of unit treatment effects for an additive model).

## Estimate the marginal structural OR — binary outcome

The marginal structural OR is a GPS-weighted logistic regression *without*
covariates in the outcome model.  It directly estimates the population-average
odds ratio comparing the two treatment levels.

```{r msor}
ate_or <- estimate_ate(
  gps,
  outcome   = Y_binary,
  estimand  = "OR"
  # treatment_levels uses the same 75th vs 25th percentile default
)
print(ate_or)
get_ate_table(ate_or)
```

## Dose-response function

We can visualise how the expected outcome changes continuously across the
full treatment range using `drf()`.

```{r drf, fig.alt = "Marginal dose-response curves for T1 and T2"}
drf_obj <- drf(ate_cont, vary = "marginal", n_grid = 60)
print(drf_obj)
plot(drf_obj)
```

For d = 2 we can also inspect the joint surface:

```{r drf_joint, fig.alt = "Joint dose-response surface heatmap"}
drf_joint <- drf(ate_cont, vary = "joint", n_grid = 40)
plot(drf_joint)
```
