---
title: "Handling Limit-of-Detection Censoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Handling Limit-of-Detection Censoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
set.seed(123)
```

## Overview

Many environmental and pharmacological studies measure treatments (e.g. chemical
concentrations) that fall below an instrument's limit of detection (LOD).
This vignette shows how `copulagps` handles LOD-censored treatments and
compares three estimation strategies:

1. **Substitution** — replace censored values with LOD / √2 (naive).
2. **IFAM** (`censoring_method = "standard"`) — iterative copula imputation
   with Rubin's rules combination across M chains.
3. **Uncensored-only** (`censoring_method = "uncens"`) — restrict outcome
   regression to fully-detected observations.

## Simulate censored data

```{r simulate}
library(copulagps)

n <- 500
X <- cbind(
  X1 = rnorm(n), X2 = rnorm(n), X3 = rnorm(n),
  X4 = rbinom(n, 1, 0.5), X5 = rbinom(n, 1, 0.5)
)

sim <- simulate_treatments(
  n             = n,
  copula_family = "clayton",
  tau           = 0.5,
  covariate_mat = X,
  beta_list     = list(c(0.4, 0, 0, 0.4, 0), c(0.4, 0, 0, -0.2, 0))
)
T_true <- sim$y   # true (unobserved) treatments

# LOD thresholds: T1 at ~20th pct, T2 at ~30th pct
lod <- c(
  T1 = quantile(T_true[, 1], 0.20),
  T2 = quantile(T_true[, 2], 0.30)
)

# Apply censoring: replace with LOD where treatment is below threshold
T_obs <- T_true
T_obs[T_true[, 1] <= lod[1], 1] <- lod[1]
T_obs[T_true[, 2] <= lod[2], 2] <- lod[2]

cat("Censoring rates: T1 =", mean(T_true[, 1] <= lod[1]) * 100, "%,",
    "T2 =", mean(T_true[, 2] <= lod[2]) * 100, "%\n")

# Generate continuous outcome from TRUE treatments (bias check)
Y <- T_true[, 1] + T_true[, 2] + 0.5 * X[, 1] + rnorm(n, 0, 1.5)
```

## Method 1: Substitution (LOD / √2)

The simplest approach is to replace all censored values with LOD / √2.
This introduces systematic measurement error and biases the ATE estimate
downward.

```{r substitution}
gps_sub <- estimate_gps(y = T_obs, x = X, lod = lod)

ate_sub <- estimate_ate(
  gps_sub, outcome = Y, covariates = X,
  censoring_method = "substitution",  # default: LOD / sqrt(2)
  sub_values       = "lod_sqrt2"
)
cat("Substitution ATE:", round(ate_sub$ate_estimates[[1]]$ate, 3),
    "(true = 2.0)\n")
```

## Method 2: IFAM — full iterative copula imputation

`estimate_gps()` with `lod` specified runs Tobit MLE for the marginals and
iterative copula imputation for the censored cells.  We run M = 5 chains and
combine estimates via Rubin's rules.

```{r ifam}
gps_ifam <- estimate_gps(
  y   = T_obs,
  x   = X,
  lod = lod,
  copula_family = "auto"
)

result_ifam <- run_chains(
  gps_obj = gps_ifam,
  outcome = Y,
  M       = 5,
  covariates = X
)

combined <- combine_rubin(result_ifam)
cat("IFAM ATE:", round(combined$pooled_ate, 3), "(true = 2.0)\n")
cat("95% CI: [", round(combined$ci_lower, 3), ",",
    round(combined$ci_upper, 3), "]\n")
```

## Method 3: Uncensored-only outcome regression

When censoring rates are modest, we can estimate the GPS from the full
sample (which correctly accounts for censoring via Tobit) but fit the
outcome model only on fully-detected observations.

```{r uncens}
ate_uncens <- estimate_ate(
  gps_ifam, outcome = Y, covariates = X,
  censoring_method = "uncens"
)
cat("Uncens ATE:", round(ate_uncens$ate_estimates[[1]]$ate, 3),
    "(true = 2.0)\n")
```

## Comparison

```{r compare}
results <- data.frame(
  Method = c("Substitution (LOD/sqrt(2))", "IFAM (M=5)", "Uncens-only"),
  ATE    = c(
    ate_sub$ate_estimates[[1]]$ate,
    combined$pooled_ate,
    ate_uncens$ate_estimates[[1]]$ate
  ),
  SE     = c(
    ate_sub$ate_estimates[[1]]$se,
    combined$pooled_se,
    ate_uncens$ate_estimates[[1]]$se
  )
)
results$Bias <- results$ATE - 2.0
print(round(results, 4))
```

The IFAM approach recovers the true ATE with the least bias.  Substitution
consistently underestimates the ATE because it replaces low (informative)
values with a single constant.  The uncensored-only estimator is unbiased
but has higher variance due to the reduced effective sample.

## Binary outcome

The same three methods work for binary outcomes by passing a binary `outcome`
vector and letting `model_type = "auto"` detect logistic regression.

```{r binary}
Y_bin <- rbinom(n, 1, plogis(-3 + T_true[, 1] + T_true[, 2] + 0.5 * X[, 1]))

ate_bin_ifam <- estimate_ate(
  gps_ifam, outcome = Y_bin, covariates = X,
  censoring_method = "standard"
)
cat("Binary ATE (risk difference):",
    round(ate_bin_ifam$ate_estimates[[1]]$ate, 4), "\n")

# Marginal structural OR (no covariates in outcome model)
ate_bin_or <- estimate_ate(
  gps_ifam, outcome = Y_bin, estimand = "OR",
  censoring_method = "standard"
)
cat("Marginal OR:", round(ate_bin_or$or_estimates[[1]]$or, 3), "\n")
```
