---
title: "Flexible Outcome Models: GAM and BART"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flexible Outcome Models: GAM and BART}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
set.seed(7)
```

## Overview

When the true outcome surface is nonlinear — for example, when the two
treatments interact — a linear outcome model is misspecified and will produce
biased ATE estimates even with perfect GPS weighting.  This vignette
demonstrates the use of GAM and BART outcome models to capture nonlinearity.

## Simulate data with an interaction surface

```{r simulate}
library(copulagps)

n <- 500
X <- cbind(
  X1 = rnorm(n), X2 = rnorm(n), X3 = rnorm(n),
  X4 = rbinom(n, 1, 0.5), X5 = rbinom(n, 1, 0.5)
)

sim <- simulate_treatments(
  n             = n,
  copula_family = "clayton",
  tau           = 0.5,
  covariate_mat = X,
  beta_list     = list(c(0.4, 0, 0, 0.4, 0), c(0.4, 0, 0, -0.2, 0))
)
T_mat <- sim$y

# Interaction outcome: true ATE (75th vs 25th pct) depends on both treatments
t_h <- apply(T_mat, 2, quantile, probs = 0.75)
t_l <- apply(T_mat, 2, quantile, probs = 0.25)

# True ATE under interaction model: sum(t_h - t_l) + 0.5*(t_h1*t_h2 - t_l1*t_l2)
true_ate_int <- sum(t_h - t_l) + 0.5 * (t_h[1] * t_h[2] - t_l[1] * t_l[2])
cat("True ATE (interaction):", round(true_ate_int, 3), "\n")

Y <- T_mat[, 1] + T_mat[, 2] + 0.5 * T_mat[, 1] * T_mat[, 2] +
     0.5 * X[, 1] + rnorm(n, 0, 1.5)
```

## Estimate GPS

```{r gps}
gps <- estimate_gps(y = T_mat, x = X, copula_family = "auto")
```

## Model 1: Linear (misspecified)

```{r lm}
ate_lm <- estimate_ate(gps, outcome = Y, covariates = X, model_type = "lm")
cat("Linear ATE:", round(ate_lm$ate_estimates[[1]]$ate, 3),
    " (true:", round(true_ate_int, 3), ")\n")
```

The linear model misses the interaction term and will be biased.

## Model 2: GAM with smooth terms

```{r gam, message = FALSE}
ate_gam <- estimate_ate(
  gps,
  outcome    = Y,
  covariates = X,
  model_type = "gam"
)
cat("GAM ATE:", round(ate_gam$ate_estimates[[1]]$ate, 3),
    " (true:", round(true_ate_int, 3), ")\n")
```

## Model 3: BART

BART uses GPS weights via an AIPW correction rather than direct weighting.
It captures complex nonlinear and interaction effects automatically.

```{r bart, eval = requireNamespace("dbarts", quietly = TRUE)}
ate_bart <- estimate_ate(
  gps,
  outcome    = Y,
  covariates = X,
  model_type = "bart"
)
cat("BART ATE:", round(ate_bart$ate_estimates[[1]]$ate, 3),
    " (true:", round(true_ate_int, 3), ")\n")
```

## Comparing DRF surfaces

The dose-response surface reveals the interaction: the effect of T1 is
larger at high values of T2, and vice versa.

```{r drf_compare, fig.alt = "Marginal DRF: linear vs GAM"}
drf_lm  <- drf(ate_lm,  vary = "marginal")
drf_gam <- drf(ate_gam, vary = "marginal")

# Stack for a side-by-side comparison
library(ggplot2)
drf_lm_df  <- do.call(rbind, drf_lm$slices);  drf_lm_df$Model  <- "Linear"
drf_gam_df <- do.call(rbind, drf_gam$slices); drf_gam_df$Model <- "GAM"
drf_comb   <- rbind(drf_lm_df, drf_gam_df)

ggplot(drf_comb, aes(x = t, y = mu_hat, colour = Model)) +
  geom_line(linewidth = 0.9) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = Model),
              alpha = 0.15, colour = NA) +
  facet_wrap(~ treatment, scales = "free_x") +
  scale_colour_manual(values = c(Linear = "#d73027", GAM = "#1a9850")) +
  scale_fill_manual(values   = c(Linear = "#d73027", GAM = "#1a9850")) +
  labs(x = "Treatment value", y = expression(hat(mu)(t)),
       title = "Dose-Response: Linear vs GAM") +
  theme_bw()
```

The GAM curves are non-straight, capturing the marginal curvature introduced
by the interaction term.

## Guidance on model selection

- **Linear model** (`model_type = "lm"`): suitable when the outcome–treatment
  relationship is approximately linear after conditioning on covariates.
- **GAM** (`model_type = "gam"`): recommended when you suspect smooth
  nonlinearity.  The number of basis functions is chosen automatically.
- **BART** (`model_type = "bart"`): the most flexible option; use when
  interactions and complex nonlinearities are expected but not known in advance.
  BART has higher computational cost and returns posterior-based SEs rather
  than sandwich SEs.
