---
title: "Custom Outcome Models: Log-Linear and Count Outcomes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Outcome Models: Log-Linear and Count Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
set.seed(99)
```

## Overview

The built-in `model_type` options (`"lm"`, `"glm"`, `"gam"`, `"bart"`) cover
the most common outcome types.  For other outcome distributions — Poisson count
data, negative binomial, Gamma responses, log-linear models — `copulagps`
provides `apply_augmented()`, which hands the GPS-augmented dataset to any
user-supplied function.

## Simulate count data

We generate a Poisson-distributed outcome where the mean depends on both
treatments and a covariate.

```{r simulate}
library(copulagps)

n <- 500
X <- cbind(
  X1 = rnorm(n), X2 = rnorm(n), X3 = rnorm(n),
  X4 = rbinom(n, 1, 0.5), X5 = rbinom(n, 1, 0.5)
)

sim <- simulate_treatments(
  n             = n,
  copula_family = "clayton",
  tau           = 0.5,
  covariate_mat = X,
  beta_list     = list(c(0.4, 0, 0, 0.4, 0), c(0.4, 0, 0, -0.2, 0))
)
T_mat <- sim$y

# Poisson log-linear outcome: log(E[Y]) = 0.3*T1 + 0.3*T2 + 0.2*X1
lambda <- exp(0.3 * T_mat[, 1] + 0.3 * T_mat[, 2] + 0.2 * X[, 1])
Y_count <- rpois(n, lambda)
cat("Mean count:", round(mean(Y_count), 2),
    "  Variance:", round(var(Y_count), 2), "\n")
```

## Estimate GPS

```{r gps}
gps <- estimate_gps(y = T_mat, x = X, copula_family = "auto")
```

## Approach 1: Use built-in `glm` with Poisson family

`estimate_ate()` supports any GLM family via the `family` argument.

```{r glm_poisson}
ate_pois <- estimate_ate(
  gps,
  outcome    = Y_count,
  covariates = X,
  model_type = "glm",
  family     = poisson(link = "log")
)
cat("Poisson GLM ATE (count scale):",
    round(ate_pois$ate_estimates[[1]]$ate, 3), "\n")
```

For count outcomes the ATE is on the absolute count scale (expected count
difference at the high vs low treatment level).

## Approach 2: Log-linear model via `apply_augmented()`

`apply_augmented()` gives you full control over the outcome model.  Your
function receives a data frame with columns `T1`, `T2`, `w` (GPS weights),
and `outcome`.

```{r apply_augmented}
my_loglinear <- function(data) {
  d_active <- data[!is.na(data$w), ]
  glm(outcome ~ T1 + T2 + X1,
      data    = d_active,
      weights = d_active$w,
      family  = gaussian(link = "log"))
}

# Attach covariates to the augmented data before passing
augmented <- get_augmented_data(gps, outcome = Y_count)
augmented$X1 <- X[, 1]

loglin_fit <- my_loglinear(augmented)
summary(loglin_fit)$coefficients[1:4, ]
```

To compute an ATE from this custom fit, use G-computation manually:

```{r gcomp_custom}
t_h <- apply(T_mat, 2, quantile, probs = 0.75)
t_l <- apply(T_mat, 2, quantile, probs = 0.25)

# Prediction data frames at t_h and t_l (all X values)
make_pdata <- function(treat) {
  df <- as.data.frame(matrix(rep(treat, n), n, byrow = TRUE))
  colnames(df) <- c("T1", "T2"); df$X1 <- X[, 1]; df
}

mu_h <- mean(predict(loglin_fit, newdata = make_pdata(t_h), type = "response"))
mu_l <- mean(predict(loglin_fit, newdata = make_pdata(t_l), type = "response"))
cat("Log-linear ATE:", round(mu_h - mu_l, 3), "\n")
```

## Approach 3: Negative binomial for overdispersed counts

If the outcome is overdispersed (variance >> mean), a negative binomial model
is appropriate.

```{r nb, message = FALSE}
library(MASS)  # for glm.nb

my_nb <- function(data) {
  d_active <- data[!is.na(data$w), ]
  MASS::glm.nb(outcome ~ T1 + T2 + X1, data = d_active, weights = d_active$w)
}

nb_fit <- my_nb(augmented)
cat("NB dispersion theta:", round(nb_fit$theta, 3), "\n")
```

## Writing a compatible `custom_fn`

Any function passed to `apply_augmented()` must:

1. Accept a `data.frame` as its first argument.
2. Have access to columns `T1`, `T2`, ..., `Td`, `w`, and `outcome`.
3. Handle `w = NA` rows (hull-trimmed): typically filter these out first with
   `data <- data[!is.na(data$w), ]`.
4. Return whatever is useful downstream (a model fit, a scalar ATE, a list).

Additional covariates can be attached to the augmented data frame before
calling `apply_augmented()`, as shown above.
